Title

### 2015-07-23 main.tex
lousy first draft

This growth in data poses significant challenges for 
researchers~\cite{marx2013biology}.
Currently, many omics applications require us to store, access, and analyze
large libraries of data.
One approach to solving these challenges is to embrace cloud computing.
Google, Inc. and the Broad Institute have teamed up to bring the GATK (Genome 
Analysis Toolkit) to the Google cloud (https://cloud.google.com/genomics/gatk).
ALSO MENTION AMAZON AWS HERE.
However, while cloud computing frees researchers from maintaining their own
data centers, and provides cost-saving benefits when computing resources are
not needed around the clock, it is no panacea.
First, in the face of disease outbreaks such as the 2014 Ebola virus epidemic 
in West Africa, analysis resources are needed at often-remote field sites.
While it is now possible to bring sequencing equipment and limited computing 
resources to remote sites, internet connectivity is still highly constrained.
Thus, accessing cloud resources for analytics may not be possible.
Moreover, cloud computing does not truly address the problem posed by
a corpus of data that is growing faster than Moore's law, because the computer
systems that make up those cloud datacenters are still themselves bound by
improvements in semiconductor technology; they still follow the trajectory
suggested by Moore's law.

ALGORITHMS FOR COMING UP WITH NOVEL BIOLOGICAL INSIGHTS not FOCUS OF THIS REVIEW
STORAGE REQUIREMENTS, NETWORK TRANSMISSION, ETC.

Computer scientists routinely exploit peculiarities in the structure of data in
order to reduce time or space complexity.
In computational biology, this approach has served researchers well.
Now-classical approaches such as principal component analysis (PCA) reduce the 
dimensionality of data in order to simplify analysis and uncover salient 
features.

@@

Seeking to take advantage of sequence redundancy, \cite{loh2012compressive}
introduced \emph{compressive genomics}, an approach that relies on compressing
data in such a way that the desired computation (such as BLAST search) can be
performed in the compressed representation.
This approach

(now, redundancy and compressive) loh2012compressive

Make use of a completely different paradigm for the structure of biological data. By combining redundancy with a low fractal dimension, able to develop algorithms that scale.

Next-gen sequencing, and some applications
Slide 27 to illustrate how sequencing works (2a, 2b)
Slide 28: sequence analysis pipeline (2c)
Redundancy - 
- 1000 genomes project, flybase, wormbase
- subsequence similarity (k-mer approaches)

Quality score compression

lossy compression - quartz

Mince~\cite{patro2015data} groups similar reads (those that share a common 
substring) together into ``buckets'', allowing that common substring to now be 
removed and treated as the bucket label, so that each read in the compressed 
representation comprises only its unique differences from the bucket label.
This allows a general-purpose compressor to achieve tight compression.

@@

Another important type of biological data is gene expression, or the quantity of
a particular gene product (usually a protein) present in the cell.
Gene expression data is useful for relating genotype to phenotype, and is 
important in studying diseases including cancer.
Usually obtained through microarrays or RNA-seq technologies, expression data is
quantitative, as each gene from a sample is associated with a numeric expression
level, and high-dimensional, as many thousands of genes may be analyzed at a
time.
While expression data lends itself to cluster analysis and probabilistic 
approaches, the high dimensionality presents challenges.
Principal Component Analysis has shown promise in reducing the dimensionality
of gene expression data~\cite{raychaudhuri2000principal, alter2000singular, yeung2001principal, misra2002interactive}.
[NEED CITATIONS FOR DIFFICULTIES -- HIGH DIM MICROARRAY DATA]
Some recent work has explored other ways to exploit the high-dimensional 
structure of the data.
SPARCLE (SPArse ReCovery of Linear combinations of Expression)~\cite{prat2011recovering}
brings ideas from compressed sensing~\cite{candes2005decoding, candes2006stable, donoho2006compressed} 
to gene expression analysis.
Specifically, SPARCLE relies on the idea that a randomly-chosen set of points 
are unlikely to be close in a lower-dimensional linear subspace.
Thus, SPARCLE approximates a gene's expression profile as a linear combination
of a few other genes' profiles, and then applies supervised learning to
predict relationships between genes.

@@

As new functional annotations and protein-protein interactions are discovered,
the number and size of these networks grows.
Diffusion component analysis (DCA)~\cite{cho} amounts to a network-based 
analogue of principle component analysis; DCA reduces the dimensionality of a
network in such a way that function annotation can be performed in the reduced 
space, significantly reducing running time while still providing competitive
prediction accuracy.

Finally, in the realm of protein structure prediction, it has been 
observed~\cite{dunbrack2002rotamer} that rather than exploring the continuous 
range of
possible dihedral angles, protein structures inhabit a smaller subspace, defined
by a narrower range of angles, referred to as 
rotamers~\cite{dunbrack1993backbone}.
The TreePack algorithm~\cite{xu2006fast} takes advantage of this observation 
for the
purpose of sidechain packing (finding the lowest-energy conformation for the
sidechain of each amino acid), using tree width to reduce the search space and
simplify dependencies.

Exploiting novel structure in biological data
- What data looks like
- actual fractal dimension and metric entropy of example data sets: table
 - metagenomics
 - protein seq
 - bacterial genomes
 - protein structure
 - chemogenomics
- Runtime
- Storage
- Search as exemplar.

Sequence-based data
- cite NRG, see cablast
- cora
- masai, gem
- (include stuff about novelty in mapping over k-mer based approaches from Cora letter to editor)
- existing mappers, cite NRG paper, and pick up where it leaves off
NOT talking about assembly
- metagenomics

Beyond sequence data
- chemogenomics
- protein structure data

Conclusions and future prospect

Other methods from image processing, etc.
Beyond biological data

for other approaches that have NOT been used in biology, see discussion
in conclusion:
other approaches that haven't been used in comp bio: 
- computational geometry (Piotr Indyk)
- succinct data structures
-

sequence approaches

structure approaches

abstraction to distance functions

no coordinate system

these observations about bio data will also hold for data from wildly different sources.

be sure to cite NRG paper, for how to extract structure from high-dimensional data.

% Bibliography
\bibliographystyle{ACM-Reference-Format-Journals}
\bibliography{main}
                             % Sample .bib file with references that match those in
                             % the 'Specifications Document (V1.5)' as well containing
                             % 'legacy' bibs and bibs with 'alternate codings'.
                             % Gerry Murray - March 2012

>>>>

This growth in data poses significant challenges for 
researchers~\cite{marx2013biology}.
Currently, many omics applications require us to store, access, and analyze
large libraries of data.
One approach to solving these challenges is to embrace cloud computing.
Google, Inc. and the Broad Institute have teamed up to bring the GATK (Genome 
Analysis Toolkit) to the Google cloud (https://cloud.google.com/genomics/gatk).
Amazon Web Services are also commonly used for computational biology 
research~\cite{schatz2010cloud}.
However, while cloud computing frees researchers from maintaining their own
data centers, and provides cost-saving benefits when computing resources are
not needed around the clock, it is no panacea.
First, in the face of disease outbreaks such as the 2014 Ebola virus epidemic 
in West Africa, analysis resources are needed at often-remote field sites.
While it is now possible to bring sequencing equipment and limited computing 
resources to remote sites, internet connectivity is still highly constrained.
Thus, accessing cloud resources for analytics may not be possible.
Moreover, cloud computing does not truly address the problem posed by
a corpus of data that is growing faster than Moore's law, because the computer
systems that make up those cloud datacenters are still themselves bound by
improvements in semiconductor technology; they still follow the trajectory
suggested by Moore's law.

Computer scientists routinely exploit peculiarities in the structure of data in
order to reduce time or space complexity.
In computational biology, this approach has served researchers well.
Now-classical approaches such as principal component analysis (PCA) reduce the 
dimensionality of data in order to simplify analysis and uncover salient 
features.
This review focuses on algorithmic advances for dealing with the growth in biological data; algorithms for novel biological insights are not its focus.

@@

Seeking to take advantage of sequence redundancy, \cite{loh2012compressive}
introduced \emph{compressive genomics}, an approach that relies on compressing
data in such a way that the desired computation (such as BLAST search) can be
performed in the compressed representation.
This approach relies on a two-stage search, referred to as \emph{coarse} and \emph{fine} search.
Coarse search is performed only on the coarse, or representative, subsequences
that represent unique data.
Any coarse sequence that is within some threshold of the query is then expanded 
into all the similar sequences it represents; the fine search is then over this
(typically small) subset of the original database.
This approach provides run-time improvements to BLAST search on 
nucleotide~\cite{loh2012compressive} and protein~\cite{daniels2013compressive}
data; moreover, these run-time improvements increase as databases grow.
Also taking advantage of sequence redundancy, Mince~\cite{patro2015data} groups 
similar reads (those that share a common 
substring) together into ``buckets'', allowing that common substring to now be 
removed and treated as the bucket label, so that each read in the compressed 
representation comprises only its unique differences from the bucket label.
This allows a general-purpose compressor to achieve tight compression.
SCALCE~\cite{hach2012scalce} also relies on a ``boosting'' scheme, reordering
reads in such a way that a general-purpose compressor achieves improved
compression ratios.

Along with a genome sequence, modern sequencing systems provide a measure of
the quality of, or confidence in, the nucleotide base read at every position.
While nucleotides reside in a 4-letter alphabet, na\"ively representable in two 
bits per position, these quality scores occupy a larger range, na\"ively
requiring six or seven bits per position.
Thus, quality scores offer low-hanging fruit with respect to compression for 
storage and transmission of sequence data.
Several approaches to compressing quality scores have been 
presented~\cite{janin2013adaptive,bonfield2013compression,ochoa2013qualcomp,yu2015quality}.
Quartz~\cite{yu2015quality} takes the unusual approach of applying lossy 
compression, by bounding the likelihood that a quality score is informative.
Surprisingly, Quartz not only demonstrates improved compression over competing
approaches, but slightly improves the accuracy of downstream variant-calling.

The idea of taking advantage of redundancy in sequence data is particularly
important in the face of new sources of data.
Driven by the plummeting costs of next-generation sequencing, the 1000 Genomes Project~\cite{10002012integrated} is pursuing a broad catalog of human 
variation; instead of producing a single reference genome for a species, many
complete genomes are catalogued.
Likewise, WormBase~\cite{harris2014wormbase} and 
FlyBase~\cite{tweedie2009flybase} are cataloguing many different species and
strains of the \emph{Caenorhabditis} worm and \emph{Drosophila} fruit fly, 
respectively.
On a potentially even larger scale is the growth of metagenomic data.
Metagenomics is the study of the many genomes (bacterial, fungal, and even 
viral) that make up a particular environment.
Such an environment could be soil from a particular region (which can lead to 
the discovery of new 
antibiotics~\cite{gillespie2002isolation,forsberg2012shared}), or it could be 
the human gut, whose microbiome has been linked to human-health concerns 
including Autism Spectrum Disorder~\cite{macfabe2012short}, 
Crohn's Disease~\cite{gevers2014treatment,manichanh2006reduced}, and 
obesity~\cite{greenblum2012metagenomic}.

The continued ability to search and analyze these growing data sets hinges on
clever algorithms that take advantage of the structure of, and redundancy 
present in, the data.
Indeed, these growing data sets ``threaten to make the arising problems 
computationally infeasible.''~\cite{berger2013computational}

One approach to more quickly searching these data sets is to come up with an
indexing technique that leverages domain knowledge about the structure of the
data.
In the realm of NGS read mapping, the Genome Multitool (GEM)
mapper~\cite{marco2012gem} makes use of a Ferragina-Manzini 
index~\cite{ferragina2000opportunistic} coupled with dynamic programming in
a compressed representation, in order to prune the search space when mapping
reads to a reference genome.
Masai~\cite{siragusa2013fast} uses an ``approximate seed'' approach to indexing
the space of possible matches, likewise pruning the search space.

Another approach is to make use of a completely different paradigm for the 
structure of biological data. 
By combining redundancy with a low fractal 
dimension~\cite{yu2015entropy}, it may be possible to develop algorithms that 
scale even as the data continue to grow.
The CORA read mapper~\cite{yorukoglu2015compressive} applies a $k$-mer based 
read-compression approach with a compressive indexing of the reference genome
(referred to as a homology table).
Short seed-clustering schemes, such as those used in Masai and 
MrsFAST~\cite{hach2010mrsfast} conceptually differ from CORA in that those 
schemes aim to accelerate only the seed-to-reference matching step.
Thus, there is a subsequent seed-extension step, which is substantially more 
costly and still needs to be performed for each read and mapping individually, 
even when seeds are clustered. 
Through its k-mer based read compression model, CORA is able to accelerate and 
achieve asymptotically sublinear scaling for both the seed-matching and 
seed-extension steps within coarse-mapping, which comprises the major bulk of 
the read-mapping computation.

In the realm of metagenomic search, BLASTX~\cite{altschul1990basic} is commonly
used to translate nucleotide reads into their possible protein sequences, and
search for them in a protein database.
Recent advances in metagenomic search tools have relied on three advances over
BLASTX: indexing, alphabet reduction, and compression.
RapSearch2~\cite{zhao2012rapsearch2} relies on alphabet reduction and a 
collision-free hash table.
The alphabet reduction, as it is reversible, can be thought of as a form of
lossless compression; a 20-letter amino acid alphabet is mapped onto a smaller
alphabet, with offsets stored to recover the original sequence in the full
alphabet.
The hash-table provides an efficient index of the database to be searched.
DIAMOND~\cite{buchfink2014fast} also relies on alphabet reduction, but uses
``shaped seeds'' instead of simple $k$-mer seeds to index the database.

The recently-released MICA~\cite{yu2015entropy} demonstrates that the 
compressive-acceleration approach of 
caBLAST~\cite{loh2012compressive} and caBLASTP~\cite{daniels2013compressive} is 
largely orthogonal to alphabet-reduction and indexing approaches.
MICA applies the compressive-acceleration framework to DIAMOND, using DIAMOND
for its ``coarse search'' phase and a user's choice of DIAMOND or BLASTX for its
``fine search'' phase; it demonstrates comparable run-time gains over DIAMOND as
caBLASTP does over BLASTP.

@@

Another important type of biological data is gene expression, or the quantity of
a particular gene product (usually a protein) present in the cell.
Gene expression data is useful for relating genotype to phenotype, and is 
important in studying diseases including cancer.
Usually obtained through microarrays or RNA-seq technologies, expression data is
quantitative, as each gene from a sample is associated with a numeric expression
level, and high-dimensional, as many thousands of genes may be analyzed at a
time.
While expression data lends itself to cluster analysis and probabilistic 
approaches, the high dimensionality presents challenges~\cite{nguyen2002tumor}.
Principal Component Analysis has shown promise in reducing the dimensionality
of gene expression data~\cite{raychaudhuri2000principal,alter2000singular,yeung2001principal,misra2002interactive}.
Some recent work has explored other ways to exploit the high-dimensional 
structure of the data.
SPARCLE (SPArse ReCovery of Linear combinations of Expression)~\cite{prat2011recovering}
brings ideas from compressed sensing~\cite{candes2005decoding,candes2006stable,donoho2006compressed} 
to gene expression analysis.
Specifically, SPARCLE relies on the idea that a randomly-chosen set of points 
are unlikely to be close in a lower-dimensional linear subspace.
Thus, SPARCLE approximates a gene's expression profile as a linear combination
of a few other genes' profiles, and then applies supervised learning to
predict relationships between genes.

@@

As new functional annotations and protein-protein interactions are discovered,
the number and size of these networks grows.
Diffusion component analysis (DCA)~\cite{cho2015diffusion} amounts to a network-based 
analogue of principle component analysis; DCA reduces the dimensionality of a
network in such a way that function annotation can be performed in the reduced 
space, significantly reducing running time while still providing competitive
prediction accuracy.

In the realm of protein structure prediction, it has been 
observed~\cite{dunbrack2002rotamer} that rather than exploring the continuous 
range of
possible dihedral angles, protein structures inhabit a smaller subspace, defined
by a narrower range of angles, referred to as 
rotamers~\cite{dunbrack1993backbone}.
The TreePack algorithm~\cite{xu2006fast} takes advantage of this observation 
for the
purpose of sidechain packing (finding the lowest-energy conformation for the
sidechain of each amino acid), using tree width to reduce the search space and
simplify dependencies.

The compressive acceleration approach of \cite{loh2012compressive} has been
generalized and adapted to non-sequence spaces as well.
Chemogenomics, the computational study of small-molecule compounds and their
affect on phenotype, relies on comparing chemical graph structures to identify
similar molecules and binding sites.
However, there are once again an increasing number of such chemical compounds
to search; the NCBI's PubChem database has grown from 31 million compounds in
January 2011 to 68 million in July, 2015.
Furthermore, comparing chemical graph structures typically involves computing
the maximal common subgraph (MCS), an NP-hard problem.
One approach to this type of search is the Small Molecule Subgraph Detector 
(SMSD)~\cite{rahman2009small}, which applies one of several MCS algorithms based
on the size and complexity of the graphs in question.
Applying a compressive acceleration approach, Ammolite~\cite{yu2015entropy}
accelerates SMSD search by an average of 150x on the PubChem database.

When the structure of a protein is known but its biological function or 
evolutionary relationships are not, researchers may search for structurally
similar proteins that are better studied~\cite{gibrat1996surprising}.
Classical tools for this involve performing pairwise structural alignments to
look for geometric similarity; DALI~\cite{holm1995dali} is still widely used,
along with other aligners such as FATCAT~\cite{ye2004fatcat} and 
Matt~\cite{menke2008matt}.
FragBag~\cite{budowski2010fragbag} accelerates protein structure search by
approximating structural alignments, which can be computationally demanding,
by instead comparing the ``bag-of-words'' from each structure.
Analogous to a term-frequency vector in information retrieval, this bag-of-words
indicates the abundance of particular, short structural motifs within a protein.
Again applying a compressive acceleration approach, 
esFragBag~\cite{yu2015entropy} clusters proteins based on the cosine distance or
Euclidean distance of their bag-of-words vectors, further accelerating FragBag's
running time by an average of 10x.

\section{Conclusions and future prospects}

The explosion of biological data, largely due to technological advances such as
next-generation sequencing, presents us with challenges as well as 
opportunities.
The promise of unlocking the secrets of diseases such as cancer, obesity,
Alzheimer's, autism spectrum disorder, and many others, as well as better 
understanding the basic science of biology, relies on researchers' ability to 
analyze the growing flood of genomic, metagenomic, structural, and interactome 
data.

The approach of compressive acceleration~\cite{loh2012compressive}, and its 
demonstrated ability to scale with the metric entropy of the 
data~\cite{yu2015entropy}, while providing orthogonal benefits to many other
useful indexing techniques, is an important tool for coping with the deluge of
data.
The extension of this compressive acceleration approach to 
metagenomics~\cite{yu2015entropy}, NGS read 
mapping and chemogenomics~\cite{yorukoglu2015compressive} suggests its 
flexibility.

The field of computational biology must continue to innovate, but also to 
incorporate the best ideas from other areas of computer science.
For example, the compressive-acceleration approach bears similarity to a metric 
ball tree, first described in the database community over twenty years 
ago~\cite{uhlmann1991satisfying}.
Other ideas from image processing, computational geometry, and other areas 
outside of biology are likely to bear fruit.
It is also likely that algorithmic ideas developed within computational biology
will become useful in other fields experiencing a data deluge, such as astronomy
or social networks~\cite{stephens2015big}.

While this Review has focused on the importance of algorithms that leverage the
structure of data to cope with the increased throughput in computational 
biology, it is also essential that techniques continue to make better 
predictions and models that demonstrate improved fidelity to wet-lab 
experiments.

\begin{table}
\caption{Metric entropy ratio (ratio of clusters to entries in database) and
fractal dimension at typical search radii for four biological data sets.
NCBI's ``NR'' non-redundant protein sequence and ``NT'' non-redundant nucleotide sequence databases are from June, 2015. Protein Databank (PDB) is from July, 2015. PubChem is from October, 2013.}
\begin{tabular}{ccc}
\hline
Data set & Metric entropy ratio & Fractal dimension \\        
\hline
Nucleotide sequences (NCBI NT) & 7:1 & 1.5\\
\hline
Protein sequences (NCBI NR) & 5:1 & 1.6\\
\hline
Protein structure (PDB) & 10:1 & 2.5\\
\hline
Chemical structure (PubChem) & 11:1 & 0.2\\
\hline
\end{tabular}
\end{table}

% Bibliography
% \bibliographystyle{ACM-Reference-Format-Journals}
\bibliographystyle{abbrv}
\bibliography{main}
